---
title: "Analysis samples no genotyped vs genotyped"
author: "Ainhoa Sanchez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, echo=FALSE}
geno <- read.table("", header = TRUE) #file with the genotyped samples data
sheet <- read.csv("")
```


```{r, echo=FALSE}
#Samples that are not in the genotyped file.
sheet_samples <- as.character(sheet$Sample_Name)
geno_samples <- as.character(unique(geno$IID1))

not_geno <- setdiff(sheet_samples, geno_samples) # In sheet_samples but not in geno_samples

cat("Number of Samples in the minerva filtered sheet that are not in the genotyped file:", length(not_geno))

```

```{r, include=FALSE}
#Get the Sample_Names that are in the genotyped file.
sheet$Sample_Name <- as.character(sheet$Sample_Name)
geno$IID1 <- as.character(geno$IID1)
geno_filtered_sheet <- sheet[sheet$Sample_Name %in% geno$IID1, ]

dim(geno_filtered_sheet)
``` 

```{r, include=FALSE}
#Use geno sheet rows to make the new collumn with the genotyped info.
sheet$Genotyped <- ifelse(sheet$Sample_Name %in% geno_filtered_sheet$Sample_Name, 1, 0)
dim(sheet)

#Save the sheet with the new collumn
write.csv(sheet, "2.filtered_new_collumn.csv", row.names = FALSE)
```

**NO GENOTYPED ANALYSIS**

```{r, echo=FALSE}
design <- read.csv("")
cosan <- read.csv("")
```

```{r, echo=FALSE}
# Create a new dataset with just the participant IDs
ids <- data.frame(
  participant_id1 = substr(design$participant_id1, 14, 19),
  participant_id2 = substr(design$participant_id2, 14, 19) 
)

```

```{r, echo=FALSE}
pids <- data.frame(
  pid_cont = design$pid_cont,
  pid_case = design$pid_case,
  participant_id1=ids$participant_id1,
  participant_id2=ids$participant_id2
)
```


```{r, echo=FALSE}
# Convert participant_id1 and participant_id2 columns to character type if they are not already
design$participant_id1 <- as.character(ids$participant_id1)
design$participant_id2 <- as.character(ids$participant_id2)

# Convert Sample_Name column to character type if it is not already
sheet$Sample_Name <- as.character(sheet$Sample_Name)

# Count how many participant_id1 match with Sample_Name
id1_matches <- sum(ids$participant_id1 %in% sheet$Sample_Name)

# Count how many participant_id2 match with Sample_Name
id2_matches <- sum(ids$participant_id2 %in% sheet$Sample_Name)

# Print the results
print(paste("Number of participant_id1 matches:", id1_matches))
print(paste("Number of participant_id2 matches:", id2_matches))
```
```{r, echo=FALSE}
caco_counts <- table(sheet$CaCo)

# Print the counts
print(caco_counts)
```
All of the Samples exist on the design file and on the sheet.
id1= case
id2= control


```{r, include=FALSE}
# Step 1: Subset "sheet" dataset based on the condition of "CaCo" column
case_samples <- sheet[sheet$CaCo == "Case", ]
ctrl_samples <- sheet[sheet$CaCo == "Ctrl", ]
dim(case_samples)
dim(ctrl_samples)
# Step 2: Merge the subset of "sheet" dataset with the "pids" dataset
case_merged <- merge(case_samples, pids, by.x = "Sample_Name", by.y = "participant_id1", all.x = TRUE)
ctrl_merged <- merge(ctrl_samples, pids, by.x = "Sample_Name", by.y = "participant_id2", all.x = TRUE)

# Step 3: Select the relevant columns and rename them
case_result <- data.frame(pid = case_merged$pid_case, Sample_Name = case_merged$Sample_Name, CaCo = case_merged$CaCo)
ctrl_result <- data.frame(pid = ctrl_merged$pid_cont, Sample_Name = ctrl_merged$Sample_Name, CaCo = ctrl_merged$CaCo)

# Combine the results into a single dataset
pid_cont_in_cosan <- rbind(case_result, ctrl_result)
dim(pid_cont_in_cosan)
```
```{r, echo=FALSE}
# Perform a left join between matches and cosan
merged_matches <- merge(pid_cont_in_cosan, cosan, by.x = "pid", by.y = "pid", all.x = TRUE)

# Add pid_f and pid_m columns to the matches dataset
pid_in_cosan_with_parents <- data.frame(
  Sample_Name = merged_matches$Sample_Name,
  CaCo = merged_matches$CaCo,
  pid = merged_matches$pid,
  pid_f = merged_matches$pid_f,
  pid_m = merged_matches$pid_m
)
```

```{r, echo=FALSE, message=FALSE}
matching_pid_m <- pid_in_cosan_with_parents[pid_in_cosan_with_parents$pid_m %in% pid_in_cosan_with_parents$pid_m[duplicated(pid_in_cosan_with_parents$pid_m)], ]
dim(pid_in_cosan_with_parents)
dim(matching_pid_m)
matching_pid_m
```

Table with the related samples of all the Samples.
pid_f is the father id and pid_m is the mother id.
If both are the same the Samples are related



**ANALYSIS OF THE GENOTYPED DATA** 

For this analysis a file produced with PLINK is used.

You can find the resulting file in: 

/home/aisanra/iastar_student_projects/aisanra/data/
minerva_genPCs_and_pruned_gens_ricopili_repl/minerva_1262_relatedness.genome

```{r, include=FALSE}
data <- read.table("", header = TRUE)
```


```{r, include=FALSE}
my_dataframe <- data.frame(IID1 = data$IID1, IID2 = data$IID2, PI_HAT = data$PI_HAT)
```

```{r, bad_samples, include=FALSE}
cosan <- subset(my_dataframe, PI_HAT > 0.125, select = c("IID1", "IID2", "PI_HAT"))
rownames(cosan) <- NULL
```

```{r, get_CaCo, echo=FALSE}

cat("Your pair of samples which Pi_value is greater than 0.125 are:")
merge_and_create_CaCo <- function(cosan, sheet, IID_column) {
  # Merge the datasets based on IID_column
  merged_data <- merge(cosan, sheet, by.x = IID_column, by.y = "Sample_Name", all.x = TRUE)
  
  # Extract CaCo column
  CaCo_column <- merged_data$CaCo
  
  # If Sample_Name does not have a corresponding value in IID_column,
  # fill it with NA
  CaCo_column[is.na(merged_data[[IID_column]])] <- NA
  
  # Return the CaCo column
  return(CaCo_column)
}

# Apply the function to IID1 and IID2 columns
cosan$IID1_CaCo <- merge_and_create_CaCo(cosan, sheet, "IID1")
cosan$IID2_CaCo <- merge_and_create_CaCo(cosan, sheet, "IID2")

# View the resulting dataset
print(cosan)

```


