---
title: "Normalization and Control Pcs Adjustment"
author: "Ainhoa Sanchez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, packages, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(ChAMP)
library(sva)
library(minfi)
library(wateRmelon)
library(data.table)
library(RColorBrewer)
library(plotly)
library(pbmcapply)
```

```{r, load, include=FALSE}
nLoad <- readRDS("filtered_probes_load.RDS")
```

```{r,data_tables, echo=FALSE, message=FALSE, include=FALSE}
rgset <- nLoad$rgSet #Get the rgset from the load data.
mset <- preprocessRaw(rgset) #only the methylated and unmethylated signals.
pd <- pData(rgset) # phenotypic data
targets <- read.metharray.sheet("")
```



**DASEN NORMALIZATION**

```{r, dasen_norm, include=FALSE}
pheno <- pData(nLoad$mset)
rownames(nLoad$mset@colData) <- paste0(pheno$Slide, "_", pheno$Array) #needed to include after champ.load for dasen normalization to work in 450K array
dnorm <- dasen(nLoad$mset)
```

```{r, save_dnorm, include=FALSE}
saveRDS(dnorm, file= "dnorm.RDS")
```

```{r, include=FALSE}
#dnorm <- readRDS("dnorm.RDS")
```


```{r, heatmap, echo=FALSE, warning=FALSE, message=FALSE}
#taking out from the phenotypes the confusing or possibly related types
pheno_corr <- pheno[, !(names(pheno) %in% c("AMP_Plate", "EstSex", "Fraction_Failed_Probes", "SexCheck"))]

#Create SVD
dnorm_beta <- as.data.frame(getBeta(dnorm))
norm_svd <- champ.SVD(beta = dnorm_beta,
              rgSet=nLoad$rgSet,
              pd=pheno_corr,
              RGEffect=TRUE,
              PDFplot=TRUE,
              Rplot=TRUE,
              resultsDir="")
```
**SVD Heatmap after Dasen normalization**

```{r, distribution_dev, include=FALSE}
pdf(file = "norm_distribution.pdf",
    width = 12, 
    height = 8)
par(mfrow=c(1,2))
# Plot distributions prior to normalization for sample 1
plotBetasByType(mset[,1],main="Raw")
typeI <- getProbeInfo(mset, type = "I")[, c("Name","nCpG")]
typeII <- getProbeInfo(mset, type = "II")[, c("Name","nCpG")]
probeTypes <- rbind(typeI, typeII)
probeTypes$Type <- rep(x = c("I", "II"), times = c(nrow(typeI), nrow(typeII)))
# Now plot the distributions of the normalized data for sample 1
plotBetasByType(dnorm[,1], probeTypes = probeTypes, main="Normalized",)
dev.off()
```

```{r, echo=FALSE}
par(mfrow=c(1,2))
# Plot distributions prior to normalization for sample 1
plotBetasByType(mset[,1],main="Raw")
typeI <- getProbeInfo(mset, type = "I")[, c("Name","nCpG")]
typeII <- getProbeInfo(mset, type = "II")[, c("Name","nCpG")]
probeTypes <- rbind(typeI, typeII)
probeTypes$Type <- rep(x = c("I", "II"), times = c(nrow(typeI), nrow(typeII)))
# Now plot the distributions of the normalized data for sample 1
plotBetasByType(dnorm[,1], probeTypes = probeTypes, main="Normalized",)
```
**Distribution of the Probe Types before and after Dasen normalization**

```{r, density_after_filter, include=FALSE}
pdf(file = "norm_density_plot.pdf",
    width = 12, 
    height = 8)
# visualise what the data looks like before and after normalisation
par(mfrow=c(1,2))
dens_raw <-densityPlot(rgset, sampGroups=targets$Sample_Group,main="Raw", legend=FALSE)
legend("top", legend = levels(factor(targets$Sample_Group)),
       text.col=brewer.pal(8,"Dark2"))
dens_norm <- densityPlot(dnorm, sampGroups=targets$Sample_Group,
            main="Normalized", legend=FALSE)
legend("top", legend = levels(factor(targets$Sample_Group)),
       text.col=brewer.pal(8,"Dark2"))
dev.off()
```

```{r, echo=FALSE}
par(mfrow=c(1,2))
dens_raw <-densityPlot(rgset, sampGroups=targets$Sample_Group,main="Raw", legend=FALSE)
legend("top", legend = levels(factor(targets$Sample_Group)),
       text.col=brewer.pal(8,"Dark2"))
dens_norm <- densityPlot(dnorm, sampGroups=targets$Sample_Group,
            main="Normalized", legend=FALSE)
legend("top", legend = levels(factor(targets$Sample_Group)),
       text.col=brewer.pal(8,"Dark2"))
```
**Comparison between the Sample densities before and after the normalization**

```{r, mds_pdf, include=FALSE}
pdf(file = "norm_mds_plot.pdf",
    width = 12, 
    height = 8)
pal <- brewer.pal(8,"Dark2")
plotMDS(dnorm_beta,pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)])
legend("top", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       bg="white", cex=0.7)
dev.off()
```

```{r, mds_plot, echo=FALSE}
plotMDS(dnorm_beta,pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)])
legend("top", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       bg="white", cex=0.7)
```
**MDS plot of the dasen normalized data**


**Adjust data for PCs from control probes**

Code provided by my supervisor.


```{r, include=FALSE}
#taking out from the phenotypes the confusing or possibly related types
pd_filt <- pheno[, !(names(pheno) %in% c("AMP_Plate", "EstSex", "Fraction_Failed_Probes", "SexCheck"))]
```





```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
pcs5_svd_ <- champ.SVD(beta = as.data.frame(5ctrlpcs),
              rgSet=nLoad$rgSet,
              pd=pd_filt,
              RGEffect=TRUE,
              PDFplot=TRUE,
              Rplot=TRUE,
              resultsDir="")
```
**Adjustment with 5 Ctrl PCs SVD Heatmap**
```{r}
pdf(file = "5prob_plot.pdf",
    width = 12, 
    height = 8)
pal <- "#4169E1"
par(mfrow=c(1,2))
plotMDS(getBeta(rgset),pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
plotMDS(5ctrlpcs,pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
dev.off()
```
```{r}
plotMDS(getBeta(rgset),pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
plotMDS(5ctrlpcs,pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
```
**SVD before and after 5 Ctrl PCs adjustment**

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
pcs7_svd_ <- champ.SVD(beta = as.data.frame(7ctrlpcs),
              rgSet=nLoad$rgSet,
              pd=pd_filt,
              RGEffect=TRUE,
              PDFplot=TRUE,
              Rplot=TRUE,
              resultsDir="")
```
**Adjustment with 7 Ctrl PCs SVD Heatmap**

```{r}
pdf(file = "mds_bf_7prob_plot.pdf",
    width = 12, 
    height = 8)
pal <- "gold"
par(mfrow=c(1,2))
plotMDS(getBeta(rgset),pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
plotMDS(7ctrlpcs,pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
dev.off()
```
```{r}
plotMDS(getBeta(rgset),pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
plotMDS(7ctrlpcs,pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
```

**SVD before and after 7 Ctrl PCs adjustment**

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
pcs8_svd_ <- champ.SVD(beta = as.data.frame(8ctrlpcs),
              rgSet=nLoad$rgSet,
              pd=pd_filt,
              RGEffect=TRUE,
              PDFplot=TRUE,
              Rplot=TRUE,
              resultsDir="")
```
**Adjustment with 8 Ctrl PCs SVD Heatmap**
```{r}
pdf(file = "mds_bf_8prob_plot.pdf",
    width = 12, 
    height = 8)
pal <- "red1"
par(mfrow=c(1,2))
plotMDS(getBeta(rgset),pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
plotMDS(8ctrlpcs,pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
dev.off()
```

```{r}
plotMDS(getBeta(rgset),pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
plotMDS(8ctrlpcs,pch = 19:19, top=1000, gene.selection="common", 
        col=pal[factor(targets$Sample_Group)], dim=c(1,2))
legend("topright", legend=levels(factor(targets$Sample_Group)), text.col=pal,
       cex=0.7, bg="white")
```
**SVD before and after 8 Ctrl PCs adjustment**
